T gates, temporary Toffolis, and halving the cost of quantum computation

### 2017-08-23 paper.tex
Investing A states

\begin{abstract}
A ``temporary Toffoli" is a doubly-controlled NOT operation that will later be uncomputed by a second double-controlled NOT.
We construct temporary Toffoli gates using four T gates; half as many as previously reported.
We demonstrate how to apply our construction (and other optimizations) to common operations and algorithms, and show significant improvement over previously reported T-costs.
For example we cut the reported T-cost of addition by 50\%, the reported T-cost of controlled-addition by 70\%, and the reported T-cost of Shor's algorithm by ?????\%.

@@

One of the downsides of the surface code is that it has no cheap mechanism to apply non-Clifford operations such as $T$ gates.
Instead, $T$ gates are performed by distilling and consuming $|A\rangle = \frac{1}{\sqrt{2}} (|0\rangle + e^{i \pi/4} |1\rangle$ states.
Consuming an $|A\rangle$ state to perform a T gate is simple (see \autoref{fig:t-equals-a}), but distilling $|A\rangle$ states has significant cost.

@@

In [[[[cite]]]] the $|A\rangle$ state distillation is done by preparing an EPR pair $(P, Q)$, moving $P$ out of the surface code and into a different error correcting code that can perform $T$ gates, and using a $T$ gate within that code to rotate and measure $P$ along an axis correlated with the diagonal $X+Y$ axis of $Q$.
If this process succeeds without an error being detected, it steers \cite{Wiseman2007} $Q$ into the $|A\rangle$ state or else into the complementary state $Z|A\rangle$.
The $Z$ rotation is easily corrected, and the result of measuring $P$ indicates whether the correction is needed or not.

The resulting `foothold' $|A\rangle$ states have a non-negligible amount of noise, but can be used to power T gates for rounds of distillation occurring entirely within the surface code.
Each round of distillation reduces the amount of noise in the final states, but at significant cost.
After accounting for distilling the $|A\rangle$ state, the spacetime volume cost of a T gate is one to two orders of magnitude higher than Clifford operations native to the surface code (such as the CNOT gate) [[[[cite]]]].

Because T gates are so expensive for the surface code, and the surface code is a likely component of future quantum computers, it is important to consider and optimize the number of T gates used by quantum circuits.
In this paper, we focus on optimizing the number of T gates needed to perform Toffoli gates that will later be uncomputed by a second Toffoli gate (``temporary Toffoli gates").
Temporary Toffolis show up in many circuits, especially ones that involve computing and uncomputing classical functions, so these optimizations are widely applicable.

Our paper is divided into six sections.
In \autoref{sec:introduction}, we motivate the problem and insert self-referential descriptions.
\autoref{sec:review} discusses how previous work on optimizing the T gate counts of Toffolis improved the cost of temporary Toffolis from 14 to 8.
In \autoref{sec:invest}, we show how to improve the cost of temporary Toffolis from 8 to 6 by investing and later recovering an $|A\rangle$ state instead of consuming it to perform T gates.
\autoref{sec:temporary-and} explains an ancilla erasing trick that improves the cost of temporary Toffolis further, from 6 to 4.
This leads to \autoref{sec:circuit-constructions}, where we demonstrate how to use our constructions to significantly improve the T-cost of several basic arithmetic tasks (e.g. halving the T-cost of addition).
Finally, \autoref{sec:conclusion} concludes and discusses future work.

@@

It is not common for adjacent Toffolis to have the same controls, but it is common for a Toffoli to later be uncomputed by a second matching Toffoli (i.e. for the Toffoli's effect to be temporary).
When this occurs, the three $T$ gates on the control qubits of the textbook construction can be omitted.
This introduces phase errors (see \autoref{fig:bad-phase-toffoli}), but the second Toffoli gate will uncompute those errors while uncomputing the state permutation (see \autoref{fig:cancelled-bad-phase-toffoli}) \cite{Barenco1995}.
This technique reduces the T-cost of paired Toffolis from 7 to 4.

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &\ctrl{1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{1}&\qw && &&&\ctrl{1}     &\ctrl{1}  &\qw &\qw &\qw    &\qw &\qw     &\ctrl{1}&\ctrl{1}     &\qw & \\
      &\ctrl{1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{1}&\qw &&=&&&\qS          &\ctrl{1}  &\qw &\qw &\qw    &\qw &\qw     &\ctrl{1}&\qSi         &\qw & \\
      &\targ   &\qw &    &\ldots &    &    &\targ   &\qw && &&&\gate{Z}\qwx &\targ     &\qw &    &\ldots &    &        &\targ   &\gate{Z}\qwx &\qw & \\
    }
  }
  \caption{
	When two Toffolis form a compute/uncompute pair, they cancel phase errors in underlying constructions.
	Fixes the problem with the construction in \autoref{fig:bad-phase-toffoli}, achieving a per-Toffoli T-cost of 4 for paired Toffolis \cite{Barenco1995}.
  }
  \label{fig:cancelled-bad-phase-toffoli}
\end{figure}

In \cite{Jones2013}, Jones shows how to perform an Toffoli with a T-cost of 4 even when not pairs.
They do so by involving an ancilla qubit and an erasure technique that turns what would have been a quantum control into a classical control.
We show this construction in \autoref{fig:jones-toffoli}.
Jones also shows how to rearrange all of the T gates into the same column, so that the Toffoli construction has a T-depth of 1 (see figure?).

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &\ctrl{1} &\qw & &   & & &     &\qw &\qw &\qw      &\qw  &\ctrl{3} &\qw &\qw      &\qw  &\qw &\qw  &\qw       &\qw &\qw    &\ctrl{1}       &\qw \\
      &\ctrl{1} &\qw & & = & & &     &\qw &\qw &\ctrl{2} &\qw  &\qw      &\qw &\ctrl{2} &\qw  &\qw &\qw  &\qw       &\qw &\qw    &\gate{Z}       &\qw \\
      &\targ    &\qw & &   & & &     &\qw &\qw &\qw      &\qw  &\qw      &\qw &\qw      &\qw  &\qw &\qw  &\targ     &\qw &\qw    &\qw\cwx        &\qw \\
      &         &    & &   & & & \qO &\qH &\qT &\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\qH &\qSi &\ctrl{-1} &\qH &\meter &\cw\cwx\bullet & \\
    }
  }
  \caption{
	Toffoli construction with 4 T gates based on measurement teleportation.
  }
  \label{fig:jones-toffoli}
\end{figure}

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &\ctrl{1} &\qw & &       &&&     &\ctrl{1} &\ctrl{1} &\qw  &&&   & & & \lstick{x} &\qw      &\qw  &\ctrl{3} &\qw &\qw      &\qw  &\qw &\qw       &\qw &\rstick{x} \\
      &\ctrl{1} &\qw & & \cong &&&     &\ctrl{2} &\qS      &\qw  &&& = & & & \lstick{y} &\ctrl{2} &\qw  &\qw      &\qw &\ctrl{2} &\qw  &\qw &\qw       &\qw &\rstick{y} \\
      &\targ    &\qw & &       &&&     &\targ    &\qw      &\qw  &&&   & & & \lstick{t} &\qw      &\qw  &\qw      &\qw &\qw      &\qw  &\qw &\targ     &\qw &\rstick{t \oplus (x \land y)} \\
      &         &    & &       &&& \qO &\targ    &\qw      &\qw  &&&   & & & \qA        &\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\qH &\ctrl{-1} &\qw &\rstick{x \land y} \\
    }
  }
  \caption{
	Our temporary Toffoli construction, which uses an $|A\rangle$ ancilla and 3 T gates.
	The $|A\rangle$ state is recovered when uncomputing the temporary Toffoli.
  }
  \label{fig:ancilla-temporary-toffoli}
\end{figure}

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &x  &&\ctrl{1} &\qw &x && && \qw  &\qw &\ctrl{2} &\qw &\qw      &\qw  &\ctrl{2} &\qw &\qw      &\qw &           \\
      &y  &&\ctrl{1} &\qw &y &&=&& \qw  &\qw &\qw      &\qw &\ctrl{1} &\qw  &\qw      &\qw &\ctrl{1} &\qw &           \\
      &xy &&\qw      &    &  && && \qSi &\qH &\targ    &\qT &\targ    &\qTi &\targ    &\qT &\targ    &\qw &|A \rangle \\
    }
  }
  \caption{
	A non-optimal unmerging circuit that uses three $T$ gates, but recovers an $|A\rangle$ state.
	Avoids measurement, but has a net T-cost of two (which is worse than \autoref{fig:unmerge-free}'s T-cost of zero).
  }
  \label{fig:unmerge-recover-a}
\end{figure}

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &x &&\ctrl{1} &\qw & x  && &&          &&\qw      &\qw  &\ctrl{2} &\qw &\qw      &\qw  &\ctrl{2} &\qw &\qw &\qw &\\
      &y &&\ctrl{1} &\qw & y  &&=&&          &&\ctrl{1} &\qw  &\qw      &\qw &\ctrl{1} &\qw  &\qw      &\qw &\qw &\qw &\\
      &  &&         &\qw & xy && &&|A\rangle &&\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\targ    &\qH &\qS &\qw &\\
    }
  }
  \caption{
	Control merging with T-cost of 4 (three T gates, one $|A\rangle$ state).
	The $|A\rangle$ state is fed directly into the circuit, instead of using an $H$ and $T$ gate on a $|0\rangle$ state (which would consume an $|A\rangle$ state anyways).
	The input $|A\rangle$ state can be recovered by using the un-merging construction in \autoref{fig:unmerge-recover-a}, reducing the net T cost of this circuit to 3.
	However, the un-merging construction in \autoref{fig:unmerge-free} is a better choice, despite not recovering the $|A\rangle$ state, because the total T-cost of merge+unmerge using \autoref{fig:unmerge-free} is 4+0=4 instead of 4+3-1=6.
  }
  \label{fig:logical-and}
\end{figure}

>>>>

\begin{abstract}
A ``temporary Toffoli" is a doubly-controlled-not operation that will later be uncomputed by a second double-controlled-not.
We construct temporary Toffoli gates using four T gates; half as many as previously reported.
We demonstrate how to apply our construction (and other optimizations) to common operations and algorithms, and show significant improvement over previously reported T-costs.
We reduce the reported T-cost of addition by 50\% and the reported T-cost of Shor's algorithm by ?????\%.

@@

One of the downsides of the surface code is that it has no cheap mechanism to apply non-Clifford operations such as $T$ gates.
Instead, $T$ gates are performed by distilling and consuming $|A\rangle = \frac{1}{\sqrt{2}} (|0\rangle + e^{i \pi/4} |1\rangle)$ states.
Consuming an $|A\rangle$ state to perform a T gate is simple (see \autoref{fig:t-equals-a}), but distilling $|A\rangle$ states has significant cost.

@@

In [[[[cite]]]] the $|A\rangle$ state distillation is done by preparing an EPR pair $(P, Q)$, translating $P$ out of the surface code and into a different error correcting code that can perform $T$ gates, and using a $T$ gate within that code to rotate and measure $P$ along an axis correlated with the diagonal $X+Y$ axis of $Q$.
If this process succeeds without an error being detected, it steers \cite{Wiseman2007} $Q$ into the $|A\rangle$ state or else into the complementary state $Z|A\rangle$.
The $Z$ rotation is easily corrected, and the result of measuring $P$ indicates whether the correction is needed or not.

The resulting `foothold' $|A\rangle$ states have a non-negligible amount of noise, but can be used to perform approximate T gates powering additional rounds of distillation occurring entirely within the surface code.
After accounting for the cost of distilling a sufficiently accurate $|A\rangle$ state, the spacetime volume cost of performing a T gate is one to two orders of magnitude higher than performing operations native to the surface code (such as the CNOT gate) [[[[cite]]]].

Because T gates are so expensive for the surface code, and the surface code is a likely component of future quantum computers, it is important to consider and optimize the number of T gates used by quantum circuits.

In this paper, we focus on optimizing the number of T gates needed to perform Toffoli gates that will later be uncomputed by a second Toffoli gate (``temporary Toffoli gates").
Temporary Toffolis show up in many circuits, especially ones that involve computing and uncomputing classical functions, so these optimizations are widely applicable.

Our paper is divided into six sections.
In \autoref{sec:introduction}, we motivate the problem and insert self-referential descriptions.
\autoref{sec:review} discusses how previous work on optimizing the T gate counts of Toffolis improved the cost of temporary Toffolis from 14 to 8.
In \autoref{sec:invest}, we show how to improve the cost of temporary Toffolis from 8 to 6 by investing and later recovering an $|A\rangle$ state instead of consuming it to perform T gates.
\autoref{sec:temporary-and} explains an ancilla erasing trick that improves the cost of temporary Toffolis further, from 6 to 4.
This leads to \autoref{sec:circuit-constructions}, where we demonstrate how to use our constructions to improve the T-cost of several basic arithmetic tasks (e.g. halving the T-cost of addition).
Finally, \autoref{sec:conclusion} concludes and discusses future work.

@@

It is not common for adjacent Toffolis to have the same controls, but it is common for a Toffoli to later be uncomputed by a second matching Toffoli (i.e. for the Toffoli's effect to be temporary).
When this occurs, the three $T$ gates on the control qubits of the textbook construction can be omitted.
This introduces phase errors (see \autoref{fig:bad-phase-toffoli}), but the second Toffoli gate can uncompute those errors while uncomputing the state permutation \cite{Barenco1995} (see \autoref{fig:cancelled-bad-phase-toffoli}).

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &\ctrl{1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{1}&\qw && &&&\ctrl{1}     &\ctrl{1}  &\qw &\qw &\qw    &\qw &\qw     &\ctrl{1}&\ctrl{1}     &\qw & \\
      &\ctrl{1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{1}&\qw &&=&&&\qS          &\ctrl{1}  &\qw &\qw &\qw    &\qw &\qw     &\ctrl{1}&\qSi         &\qw & \\
      &\targ   &\qw &    &\ldots &    &    &\targ   &\qw && &&&\gate{Z}\qwx &\targ     &\qw &    &\ldots &    &        &\targ   &\gate{Z}\qwx &\qw & \\
    }
  }
  \caption{
	When two Toffolis form a compute/uncompute pair, they can cancel each others' phase errors.
	Fixes the problem with the construction in \autoref{fig:bad-phase-toffoli}, and achieves a per-Toffoli T-cost of 4 for paired Toffolis \cite{Barenco1995}.
  }
  \label{fig:cancelled-bad-phase-toffoli}
\end{figure}

The paired-phase-correction trick improves the T-cost of temporary Toffolis from 14 (seven to compute, seven to uncompute) to 8 (four to compute, four to uncompute).

In \cite{Jones2013}, Jones shows how to perform an {\em unpaired} Toffoli with four T gates.
They use an ancilla qubit and an erasure technique that turns what would have been a quantum control into a classical control.
They also re-arrange the remaining T gates into a single column, creating a circuit with T-depth of 1.
We show this construction in \autoref{fig:jones-toffoli}.

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &\ctrl{1} &\qw & &   & & &     &\qw &\qw &\qw      &\qw  &\ctrl{3} &\qw &\qw      &\qw  &\qw &\qw  &\qw       &\qw &\qw    &\ctrl{1}       &\qw \\
      &\ctrl{1} &\qw & & = & & &     &\qw &\qw &\ctrl{2} &\qw  &\qw      &\qw &\ctrl{2} &\qw  &\qw &\qw  &\qw       &\qw &\qw    &\gate{Z}       &\qw \\
      &\targ    &\qw & &   & & &     &\qw &\qw &\qw      &\qw  &\qw      &\qw &\qw      &\qw  &\qw &\qw  &\targ     &\qw &\qw    &\qw\cwx        &\qw \\
      &         &    & &   & & & \qO &\qH &\qT &\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\qH &\qSi &\ctrl{-1} &\qH &\meter &\cw\cwx\bullet & \\
    }
  }
  \caption{
	Toffoli construction with T-cost of 4 from \cite{Jones2013}.
  }
  \label{fig:jones-toffoli}
\end{figure}

@@

The seed for this paper was an idea that accidentally improved the T-cost of a Toffoli from 7 to 6 in a surprising way.
Instead of performing a Toffoli directly onto the target qubit, we performed the Toffoli indirectly via an ancilla.
We started with a clean ancilla in the $|0\rangle$ state, applied a Toffoli to store the logical-and of the two controls in the ancilla, used the ancilla to control a CNOT onto the intended target, then uncomputed the ancilla.

Initially, this indirect-Toffoli construction seems pointless.
It has a T-cost of 8 even when using the phase-neglected construction, which is worse than the naive T-cost of 7.
However, as shown in \autoref{fig:indirect-toffoli}, it turns out that the last T gate in the circuit is being used to uncompute an $|A\rangle$ state on the ancilla.
Dropping this T gate not only reduces the T-cost from 8 to 7, it recovers an $|A\rangle$ state that can be used to power some other T gate.
This improves the net T-cost to 6.

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &\ctrl{1} &\qw && &&&    &\qw &\qw &\qw      &\qw  &\ctrl{3} &\qw &\qw      &\qw  &\qw &\qw       &\qw &\qw &\qw      &\qw  &\ctrl{3} &\qw &\qw      &\qw  &\qw &\qw &&\\
      &\ctrl{1} &\qw &&=&&&    &\qw &\qw &\ctrl{2} &\qw  &\qw      &\qw &\ctrl{2} &\qw  &\qw &\qw       &\qw &\qw &\ctrl{2} &\qw  &\qw      &\qw &\ctrl{2} &\qw  &\qw &\qw &&\\
      &\targ    &\qw && &&&    &\qw &\qw &\qw      &\qw  &\qw      &\qw &\qw      &\qw  &\qw &\targ     &\qw &\qw &\qw      &\qw  &\qw      &\qw &\qw      &\qw  &\qw &\qw &&\\
      &         &    && &&&\qO &\qH &\qT &\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\qH &\ctrl{-1} &\qH &\qT &\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\qH &\qw 
          \gategroup{4}{6}{4}{10}{.7em}{--} \gategroup{4}{26}{4}{30}{.7em}{--} &|0\rangle & \\
      &         &    && &&&    &    |A\rangle &  & &     &         &    &         &     &    &          &    &    &         &    &         &     &         &     & |A\rangle
    }
  }
  \caption{
	Performing a Toffoli indirectly, by applying a pair of Toffolis to an ancilla and using the intermediate value to control a CNOT onto the actual target, appears to have a T-cost of 8.
	However, the initial T gate is being used to compute an $|A\rangle$ state and the final T gate is being used to uncompute the $|A\rangle$ state.
    \\	
	Instead of spending $|A\rangle$ states to perform T gates to make and unmake $|A\rangle$ states, we pass in and later recover a single $|A\rangle$ state.
	This ``investment" reduces the net T-cost to 6.
  }
  \label{fig:indirect-toffoli}
\end{figure}

A Toffoli with T-cost of 6 isn't optimal, but the reduction from the textbook T-cost of 7 was done in an unexpected way.
There were likely other situations where this optimization would improve on the state of the art, and we soon found one: temporary Toffolis.
When a Toffoli is going to be uncomputed, it is not necessary to compute, uncompute, recompute, and reuncompute the ancilla qubit.
Instead, compute the ancilla for the first Toffoli, keep the ancilla around to make performing the second Toffoli easy, then uncompute the ancilla.
Separating the control-merging step from the control-unmerging step improves the T-cost of temporary Toffolis from 8 to 6.

[[[[Keeping ancilla around is counter-intuitive, because the ancilla represents entangled garbage that can prevent interference from occurring correctly.
It just so happens that it makes sense to do so for uncomputation tasks related to classical functions.]]]]

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &\ctrl{1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{1}&\qw && &&&    &\qw     &\qw  &\ctrl{3}&\qw &\qw     &\qw  &\qw &\qw      &\qw &\qw &\qw    &\qw &\qw &\qw      &\qw &\qw &\qw     &\qw  &\ctrl{3}&\qw &\qw     &\qw &&\\
      &\ctrl{1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{1}&\qw &&=&&&    &\ctrl{2}&\qw  &\qw     &\qw &\ctrl{2}&\qw  &\qw &\qw      &\qw &\qw &\qw    &\qw &\qw &\qw      &\qw &\qw &\ctrl{2}&\qw  &\qw     &\qw &\ctrl{2}&\qw &&\\
      &\targ   &\qw &    &\ldots &    &    &\targ   &\qw && &&&    &\qw     &\qw  &\qw     &\qw &\qw     &\qw  &\qw &\targ    &\qw &    &\ldots &    &    &\targ    &\qw &\qw &\qw     &\qw  &\qw     &\qw &\qw     &\qw &&\\
      &        &    &    &       &    &    &        &    && &&&\qA &\targ   &\qTi &\targ   &\qT &\targ   &\qTi &\qH &\ctrl{-1}&\qw &\qw &\qw    &\qw &\qw &\ctrl{-1}&\qH &\qT &\targ   &\qTi &\targ   &\qT &\targ   &\qw &|A\rangle &\\
    }
  }
  \caption{
	A temporary Toffoli construction with a net T-cost of 6.
	Invests an $|A\rangle$ state when computing the Toffoli, holds an ancilla qubit storing the logical-and of the controls until it's time to uncompute the Toffoli, then recovers the $|A\rangle$ state.
  }
  \label{fig:ancilla-temporary-toffoli}
\end{figure}

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=1.5em @C=0.7em {
      &x &&\ctrl{1} &\qw & x  && &&          &&\qw      &\qw  &\ctrl{2} &\qw &\qw      &\qw  &\qw &\qw &\qw &\\
      &y &&\ctrl{1} &\qw & y  &&=&&          &&\ctrl{1} &\qw  &\qw      &\qw &\ctrl{1} &\qw  &\qw &\qw &\qw &\\
      &  &&         &\qw & xy && &&|A\rangle &&\targ    &\qTi &\targ    &\qT &\targ    &\qTi &\qH &\qS &\qw &\\
    }
  }
  \caption{
	Control merging with T-cost of 4 (three T gates, one $|A\rangle$ state).
	The $|A\rangle$ state is fed directly into the circuit, instead of using an $H$ and $T$ gate on a $|0\rangle$ state (which would consume an $|A\rangle$ state anyways).
	The input $|A\rangle$ state can be recovered by using the un-merging construction in \autoref{fig:unmerge-recover-a}, reducing the net T cost of this circuit to 3.
	However, the un-merging construction in \autoref{fig:unmerge-free} is a better choice, despite not recovering the $|A\rangle$ state, because the total T-cost of merge+unmerge using \autoref{fig:unmerge-free} is 4+0=4 instead of 4+3-1=6.
  }
  \label{fig:logical-and}
\end{figure}
